Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-27 submission/manuscript.Rmd
begin CGI discussion edits

Unlike the previously characterized germline gene suppressors that place repressive histone modifications[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes the active histone mark histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. Yet, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Although dispensible in ESCs, KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. In support of this, CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017]. We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at germline CpG islands. KDM5C's shifting role in germline gene repression during the transition from na&#239;ve to primed pluripotency indicates chromatin modifiers can switch between catalytic and non-catlytic gene regulatory mechanisms at the same loci over development.

>>>>

Unlike the previously characterized germline gene suppressors that place repressive histone modifications[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes the active histone mark histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. Yet, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Although dispensible in ESCs, KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. In support of this, CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017]. We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at CGIs for select germline gene promoters. KDM5C's shifting role in germline gene repression during the transition from na&#239;ve to primed pluripotency indicates the requirement of catalytic and non-catlytic gene regulatory mechanisms can change depending upon the locus and developmental stage.
